Circular Regression Trees and Forests with an Application to Probabilistic Wind Direction Forecasting

### 2019-11-14 pkg/circtree/inst/wind_paper/draft/draft.Rnw
Small revisions (Lisa and Moritz)
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2876 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 78d452c84de6c440bff6b61bad3f8ad865d95fa7

As a motivational example Figure~\ref{fig:tree_ibk} displays a distributional
tree applied to predict the wind direction at the airport of Innsbruck in the following hour. 
Innsbruck lies at the bottom of a deep valley in the Alps. Topography channels wind along the 
west-east valley axis or along a tributary valley intersecting from the south. Hence, pressure 
gradients to which valley wind regimes react both west and east of the airport are considered 
as covariates along with other meteorological measurements at the airport (lagged by one hour),
such as wind direction and wind speed at Innsbruck Airport.

@@

A more extensive case study of circular regression trees and forests applied to probabilistic 
forecasting of wind direction at the airports of Innsbruck and Vienna is presented in 
Section~\ref{sec:wind} where their performance is also compared to other commonly used models.
Before that, distributional models for circular data are discussed in Section~\ref{sec:prob_circ} 
and the methodology of circular regression trees and forest and its features are presented in 
Section~\ref{sec:tree_forest}.

@@

\subsection{Data}
For both airports, Innsbruck and Vienna, observations of various meteorological quantities 
such as wind direction, temperature, air pressure and humidity
measured either directly at the airports or at surrounding weather stations are used as 
covariates in order to predict the wind direction at the corresponding airport 
in the following hour(s).
Next to the current measurements at each of the weather stations also differences in these 
quantities between the considered stations are included. For example, differences in air 
pressure between Innsbruck and either a western or eastern location are expected to have a 
high influence on wind directions. Moreover, changes of these quantities over time are 
included as well as daytime. In total, 550 covariates are provided for Innsbruck and 522 
covariates for Vienna. 
A list of all covariates together with their abbreviations as used in this paper
can be found in Table~\ref{tab:covariates}.
The data sets each provide observations from 5 years, in particular from 2014 to 2017, both included.
After eliminating observations with at least one missing value a total number of 26861 time points
for Innsbruck and 33853 for Vienna are available.
As winds with very low speed tend to vary their direction rapidly and almost randomly they are not 
expected to have a meaningful impact on predictions. Moreover, the performance and the stability 
especially of the employed reference models might suffer from the influence of these ``noise'' 
observations. Therefore, only observations corresponding to a wind speed of at least $1$ m/s are 
considered.

@@

\subsection{Models and evaluation}\label{sec:wind:models}
For a fair evaluation of circular regression trees and forest and to
investigate whether circular trees and forests can be applied as a reasonable
alternative to already existing approaches, three additional statistical models are
applied in this study to probabilistically forecast wind directions. Two
approaches are based on existing approaches in the field of meteorological
forecasting, one approach is a state-of-the-art model to forecast circular
variables.  The next paragraphs introduce the alternative approaches and
provide for all statistical models, including the circular tree and forest, the
model specification employed in this study.

@@

In addition to the raw CRPS, to assess differences in the
improvement of the various statistical models over the climatological model
used as a reference, CRPS-based skill scores are computed according to
\begin{equation} 
  \text{CRPSS}_{\text{method}} = 1 -
\frac{\text{CRPS}_{\text{method}}}{\text{CRPS}_{\text{climatology}}}.
\end{equation}

All scores presented in the next section are out-of-sample: For the persistency
model only time points prior to the time of interest are used, for all other
models a five-fold cross-validation is employed using up to four calendar years
for model training and the remaining single calendar year for validation. Due
to the large sample size of 24-hourly values per day over five years, some kind
of temporal aggregation is needed to ensure a correct visual comparison of the
individual methods. As for the employed models the variability of the
predictive performance over the five years is lower than over a single day or
over a single year, CRPS and CRPS skill scores are aggregated over the
respective five validation years; this yields 24 hourly scores per 12 months
averaged over the five validation years.

@@

The different site characteristics between the airports Innsbruck
(Figure~\ref{fig:boxplot_crpsraw}a,\,c) and Vienna
(Figure~\ref{fig:boxplot_crpsraw}b,\,d) seem to have an effect on the absolute
level of the respective model performances and on their respective performance
variation. At Innsbruck, due to the surrounding mountains a limited number of
possible wind directions exists, namely the three wind regimes discussed for
Figure~\ref{fig:tree_ibk} in Section~\ref{sec:introduction}. Therefore, for
Innsbruck the wind direction remains rather constant in one of these possible
states, but once a change takes place it is mostly a major wind direction
shift.  Hence, due to the few wind regimes the rather inflexible climatological
and linear models score relatively well with similar values as the other models
(Figure~\ref{fig:boxplot_crpsraw}a,\,c). In addition, at Innsbruck the
potential high prediction errors in case of a paradigm shift seem to lead to
an higher variation in the predictive performance for all models in comparison
to Vienna; this variation is especially high for the persistency model due to
its strong vulnerability to abrupt wind shifts. On the contrary, at Vienna the
geographic surroundings do not constrain any wind direction which leads to
smaller and less abrupt changes in the wind direction as well as less
pronounced wind regimes. This seems to weaken the predictive performance of the
climatological and linear models, and to reduce the performance variability for
all models (Figure~\ref{fig:boxplot_crpsraw}b,\,d).

@@

\begin{figure}[t]
  \centering
  \setkeys{Gin}{width=1.0\textwidth}
  %\hspace{-2cm}
  %\includegraphics{boxplot_crpsraw.pdf}
  \includegraphics{_plot_circforest_validation_crpsraw_agg_comparison_with_lowff_v14b.pdf}
  \caption{CRPS skill scores for wind direction forecasts based on the full
    predictive von~Mises distribution for $+1$ and $+3$\,h forecasts at the
    airports Innsbruck and Vienna. Each box-and-whisker contains $24$~hourly scores
    per $12$~months averaged over the $5$~validation years; this yields a total of
    $288$~yearly mean values. The scores are shown for the reference models climatology, 
    persistence and the linear model, as well as for the circular regression tree and
    forest.}
  \label{fig:boxplot_crpsraw} 
\end{figure}

@@

\section{Summary and conclusion}
Distributional trees for circular responses are established by coupling
model-based recursive partitioning with the von Mises distribution.
The resulting trees can capture nonlinear changes, shifts, and potential interactions
in covariates without prespecification of such effects. This is particularly
useful for modeling wind direction in mountainous terrain where wind shifts
can occur due to turns of the pressure gradients along a valley.

>>>>

As a motivating example Figure~\ref{fig:tree_ibk} displays a distributional
tree applied to predict the wind direction at the airport of Innsbruck in the
following hour.  Innsbruck lies at the bottom of a deep valley in the Alps.
Topography channels wind along the west-east valley axis or along a tributary
valley intersecting from the south. Hence, pressure gradients to which valley
wind regimes react are considered as covariates along with other meteorological
measurements at the airport (lagged by one hour), such as wind direction and
wind speed at the airport.

@@

A more extensive case study of circular regression trees and forests applied to
probabilistic forecasting of wind direction at the airports of Innsbruck and
Vienna is presented in Section~\ref{sec:wind} where their performances are
benchmarked to alternative commonly used approaches.  Beforehand,
distributional models for circular data are discussed in
Section~\ref{sec:prob_circ} and the methodology of circular regression trees
and forest and its features are presented in Section~\ref{sec:tree_forest}.
After the case study, a comprehensive summary and conlusion is given in
Section~\ref{sec:summary}.

@@

\subsection{Data}
As response variables, $10$\,min mean wind direction observations are employed at
the airports Innsbruck and Vienna on an hourly temporal resolution. As
predictor variables, we use observations of various meteorological quantities
such as wind direction, temperature, air pressure and humidity directly at the
respective airports, as well as within the vicinity of the airports. For
Innsbruck, meteorological variables are observed at XXX different locations
along the intersecting valleys; whereas for Vienna, located in the flat,
meteorological variables are observed at XXX different locations within an area
of about $10$\,km$^2$. In addition, we employ temporal differences, averages, minima, maxima over the last 3\,h,

of ias well as tempat each of the weather stations also differences in these 
quantities between the considered stations are included. For example, differences in air 
pressure between Innsbruck and either a western or eastern location are expected to have a 
high influence on wind directions. Moreover, changes of these quantities over time are 
included as well as daytime. In total, 550 covariates are provided for Innsbruck and 522 
covariates for Vienna. 
A list of all covariates together with their abbreviations as used in this paper
can be found in Table~\ref{tab:covariates}.
The data sets each provide observations from 5 years, in particular from 2014 to 2017, both included.
After eliminating observations with at least one missing value a total number of 26861 time points
for Innsbruck and 33853 for Vienna are available.
As winds with very low speed tend to vary their direction rapidly and almost randomly they are not 
expected to have a meaningful impact on predictions. Moreover, the performance and the stability 
especially of the employed reference models might suffer from the influence of these ``noise'' 
observations. Therefore, only observations corresponding to a wind speed of at least $1$ m/s are 
considered.

@@

\subsection{Models and evaluation}\label{sec:wind:models}
For a fair evaluation of circular regression trees and forest and to
investigate whether they can be applied as a reasonable alternative to already
existing approaches, three additional statistical models are employed in this
study for probabilistic forecasting of wind directions. Two of them are based
on existing approaches in the field of meteorological forecasting while the
third is a state-of-the-art model to forecast circular response variables. The
next paragraphs introduce the reference methods and provide the corresponding
model specifications as employed in this study for all considered model
including the circular tree and forest.

@@

In addition to the raw CRPS, CRPS-based skill scores are computed to assess
differences in the improvement of the various statistical models over the
climatological model used as a reference:
\begin{equation} 
  \text{CRPSS}_{\text{method}} = 1 -
\frac{\text{CRPS}_{\text{method}}}{\text{CRPS}_{\text{climatology}}}.
\end{equation}

All scores presented in the next section are out-of-sample: For the persistency
model only time points prior to the time of interest are used, for all other
models a five-fold cross-validation is employed using up to four calendar years
for model training and the remaining single calendar year for validation. Due
to the large sample size of 24 hourly values per day over five years, some kind
of temporal aggregation is needed to ensure a correct visual comparison of the
individual methods. As observed for the employed models, the variability of the
predictive performance over the five years is lower than over a single day or
over a single year. Hence, CRPS and CRPS skill scores are aggregated over the
respective five validation years which yields 24 hourly scores per month
averaged over the five validation years.

@@

The different site characteristics between the airports Innsbruck
(Figure~\ref{fig:boxplot_crpsraw}a,\,c) and Vienna
(Figure~\ref{fig:boxplot_crpsraw}b,\,d) seem to have an effect on the absolute
level of the model performances and on their respective performance
variation. At Innsbruck, due to the surrounding mountains a limited number of
possible wind directions exists, namely the three wind regimes discussed for
Figure~\ref{fig:tree_ibk} in Section~\ref{sec:introduction}. Therefore, for
Innsbruck the wind direction remains rather constant in one of these possible
states, but once a change takes place it is mostly a major wind direction
shift.  Hence, due to the few wind regimes the rather inflexible climatological
and linear models score relatively well with similar values as the other models
(Figure~\ref{fig:boxplot_crpsraw}a,\,c). In addition, at Innsbruck the
potential high prediction errors in case of a paradigm shift seem to lead to
a higher variation in the predictive performance for all models in comparison
to Vienna; this variation is especially high for the persistency model due to
its strong vulnerability to abrupt wind shifts. On the contrary, at Vienna the
geographic surroundings do not constrain any wind direction which leads to
smaller and less abrupt changes in the wind direction as well as less
pronounced wind regimes. This seems to weaken the predictive performance of the
climatological and linear models, and to reduce the performance variability for
all models (Figure~\ref{fig:boxplot_crpsraw}b,\,d).

@@

\begin{figure}[t]
  \centering
  \setkeys{Gin}{width=1.0\textwidth}
  %\hspace{-2cm}
  %\includegraphics{boxplot_crpsraw.pdf}
  \includegraphics{_plot_circforest_validation_crpsraw_agg_comparison_with_lowff_v14b.pdf}
  \caption{CRPS skill scores for wind direction forecasts based on the full
    predictive von~Mises distribution for $+1$ and $+3$\,h forecasts at the
    airports Innsbruck and Vienna. Each box-and-whisker contains $24$~hourly scores
    per for each of the $12$~months averaged over the $5$~validation years which yields a total of
    $288$~yearly mean values. The scores are shown for the climatological, 
    persistency and the linear model as well as for the circular regression tree and
    forest.}
  \label{fig:boxplot_crpsraw} 
\end{figure}

@@

\section{Summary and conclusion}\label{sec:summary}
Distributional trees for circular responses are established by coupling
model-based recursive partitioning with the von Mises distribution.
The resulting trees can capture nonlinear changes, shifts, and potential interactions
in covariates without prespecification of such effects. This is particularly
useful for modeling wind direction in mountainous terrain where wind shifts
can occur due to turns of the pressure gradients along a valley.
